---
title: "Stat 542 Coding Assignment 2"
author: "Doug Mulley"
date: "9/18/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(glmnet)
library(ggplot2)
library(gridExtra)

rm(list = ls())
set.seed(9764)
```

## Part 1: BostonHousing1.data

Load `BostonHousing1.Rdata` which has 16 variables including the response variable `Y`. The data has been pre-processed, so no need to apply any transformation.

```{r}
load('BostonHousing1.Rdata')
myData = Housing1

n = nrow(myData)
p = ncol(myData) - 1

X = data.matrix(myData[,-1])
Y = myData[,1]

T = 50
```

1. Repeat  the  following  simulation  50  times.   In  each  iteration,  randomly split the data into two parts, 75% for training and 25% for testing.  fit the model based on the training data and obtain a prediction on the test data, record the mean squared prediction error (MSPE) on the test set, the selected model-size or effect dimension (for Ridge), and the computation time for each procedure.

```{r}
ntest = round(n * 0.25)
ntrain = n - ntest

all.test.id = matrix(0, nrow = ntest, ncol = T)

for (t in 1:T) {
  all.test.id[,t] = sample(1:n, ntest)
}

model_names = c("Full", 
                "AIC.F", "AIC.B", 
                "BIC.F", "BIC.B", 
                "R.min", "R.1se", 
                "L.min", "L.1se",
                "L.Refit")

nmodels = length(model_names)

mspe = matrix(0, T, nomodels)
colnames(mspe) = model_names

computation_times = matrix(0, T, nmodels)
colnames(computation_times) = model_names

calculateMSPE <- function (test.id, predictedY) {
  mean((Y[test.id] - predictedY) ^ 2)
}
```

* __Full__

```{r}
model.index = 1

for (t in 1:T) {
  time.start = proc.time()
  
  test.id = all.test.id[,t]
  full.model = lm(Y ~ ., data = myData[-test.id,])
  Ytest.pred = predict(full.model, newdata = myData[test.id,])
  
  time.span = proc.time() - time.start
  computation_times[t, model.index] = time.span["elapsed"]
  
  mspe[t, model.index] = calculateMSPE(test.id, Ytest.pred)
}
```

* __AIC - Forward__

```{r}
model.index = 2

for (t in 1:T) {
  time.start = proc.time()
  test.id = all.test.id[, t]
  full.model = lm(Y ~ ., data = myData[-test.id, ])
  stepAIC = step(lm(Y ~ 1, data = myData[-test.id, ]),
                 list(upper = full.model),
                 trace = 0,
                 direction = "forward")
  Ytest.pred = predict(stepAIC, newdata = myData[test.id, ])
  
  time.span = proc.time() - time.start
  computation_times[t, model.index] = time.span["elapsed"]
  
  mspe[t, model.index] = calculateMSPE(test.id, Ytest.pred)
}
```

* __AIC - Backward__

```{r}
model.index = 3

for (t in 1:T) {
  time.start = proc.time()
  test.id = all.test.id[, t]
  full.model = lm(Y ~ ., data = myData[-test.id, ])
  stepAIC = step(full.model,
                 trace = 0,
                 direction = "backward")
  Ytest.pred = predict(stepAIC, newdata = myData[test.id, ])
  
  time.span = proc.time() - time.start
  computation_times[t, model.index] = time.span["elapsed"]
  
  mspe[t, model.index] = calculateMSPE(test.id, Ytest.pred)
}
```

